Data Specialization for a Staged Lambda Calculus

### 2014-11-07 paper/splitting.tex
Big change to redo the transition from partial value masking to stage 1 splitting.

Given a term in the source language \lang\ where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two separate terms, each performing the
subcomputations corresponding to each stage.  More precisely, given a
\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
unstaged target language \langmono\ such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
the staged language produces the same result as evaluating~$e_1$, followed by evaluating 
$e_2$ with the result from~$e_1$.

@@

%For example, the partial value below contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
%computation $\letin \yhat {\mathtt{7+5}} \yhat$.

% such as in the example below:

%For example, consider the stage 1 term
%\begin{lstlisting}
%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
%\end{lstlisting}
%which evaluates via $\redonesym$ to a 
%
%residual table and multi-stage tuple

%This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
%computation $\letin \yhat {\mathtt{7+5}} \yhat$.

Recall that \lang\ partial values and residual tables together may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
Intuitively, splitting this term should generate an $e1$ that reduces to the stage 1 content of the partial value (3), and an $e2$ corresponding to the stage~\bbtwo\
computation $\letin \yhat {\mathtt{7+5}} \yhat$.
To precisely define the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\ that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
masking functions.
% written $\masko{{-}}$ and $\maskt{{-};{-}}$.

\emph{Stage~\bbone\ masking} of the partial value $v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
content, $\masko{v}$ replaces all $\next$ subterms of $v$ with \texttt{()}.  Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
the same structure as~$v$.

\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context $\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
requires access to the residual table $\xi$ because all stage~\bbtwo\
content in $v$ has been lifted out into $\xi$. Stage~\bbtwo\ masking discards
stage~\bbone\ content in $v$ by replacing base constants with~\texttt{()},
replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ splitting, stage~\bbtwo\
splititing masks functions by splitting their bodies as general stage~\bbone\ terms, and
returning the stage~\bbtwo\ component. After discarding stage~\bbone\ content from $v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
producing a term in \langmono.

@@

\ur{Fill this in.}

FOLLOW OUTLINE:

\begin{itemize}
\item The structure of the splitting judgment.
\item The equivalence condition.
\item Walk through the cases.
\end{itemize}

\ur{What follows is still a bit vague to me. Some
  suggestions/questions: are ``term'' and ``computation'' being used
  interchangeably? Not sure I understand this.  Isn't ``multi-stage''
  tuple a partial-value?}

@@

\ur{It starts getting rough here, mostly because combined term is not
  defined.  Specifically, we should say what
  the  form of the splitting judgment is. }

Given the definition of masking, splitting a stage \bbone\ term $\colone e A$ 
(where $e\mathbin{\redonesym}[\xi;v]$), yields a \emph{combined term}
$c$ that performs all stage~\bbone\ computations in $e$ and a \emph{resumer} function $l.r$ that represents all stage~\bbtwo\ computations.
Whereas in stage~\bbtwo\ term splitting,
$p$ constitutes only the precomputation to the resumer, in stage~\bbone\ splitting we require $c$ to reduce (via~$\redsym$) to the tuple $(y,z)$, where $y$ is an \emph{immediate result} that is identical to $\masko{\xi;v}$, and $z$ is the required precomputation such that $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.

Splitting stage~\bbone\ terms is implemented by the judgment $e \splitonesym [c,l.r]$ in \ref{fig:termSplitOne}.

% The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
% resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
% which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
% \texttt{((),12)}, which agrees with the stage \bbtwo\ mask.

>>>>

Given a term $e$ in the source language \lang\ where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two terms, $e_1$ and $e_2$, in an unstaged target language \langmono\ 
%each performing the
%subcomputations corresponding to each stage.  More precisely, given a
%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
%unstaged target language \langmono\
such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
the staged language produces the same result as evaluating~$e_1$, followed by evaluating 
$e_2$ with the result from~$e_1$.

@@

Recall that \lang\ partial values and residual tables together may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
Intuitively, splitting this term should generate an $e_1$ that reduces to the stage 1 content of the partial value (3), and an $e_2$ corresponding to the stage~\bbtwo\
computation $\letin \yhat {\mathtt{7+5}} \yhat$.
To precisely define the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\
that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
masking functions.

\emph{Stage~\bbone\ masking} of the partial value $v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
content, $\masko{v}$ replaces all $\next$ subterms of $v$ with \texttt{()}.  Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
the same structure as~$v$.  For example, the stage~\bbone\ mask of the partial value given above is \texttt{(3,())}.

\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context $\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
requires access to the residual table $\xi$ because all stage~\bbtwo\
content in $v$ has been lifted out into $\xi$. Stage~\bbtwo\ masking discards
stage~\bbone\ content in $v$ by replacing base constants with~\texttt{()},
replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ masking, stage~\bbtwo\
masking splits the bodies of lambdas as general stage~\bbone\ terms, and
returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from $v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
producing a term in \langmono. The stage~\bbtwo\ mask of the partial value and residual table given above is:
\begin{lstlisting}
let yhat=7+5 in ((),yhat)
\end{lstlisting}

@@

@@

Splitting a stage \bbone\ term $\colone e A$ 
(where $e\mathbin{\redonesym}[\xi;v]$), yields a term 
$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that represents all stage~\bbtwo\ subcomputations.
For splitting to match the results of evaluating $e$ in \lang,
we require that $e_1$ reduce to the tuple $(y,z)$,
where $y$ is identical to $\masko{\xi;v}$, and 
$\letin l z r$ reduces to the same value as $\maskt{\xi;v}$. Intuitively, $e_1$ performs both the stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (\emph{precomputations}) required to produce the input for the resumer.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [e_1,l.r]$ in \cref{fig:termSplitOne}, proceeds
recursively on the structure of~$e$.
